Perceived value of video games, but not hours played, predicts mental well-being in adult Nintendo players

### 2024-06-26 ms.qmd
addressing latest feedback

---
title: "Well-being is linked to the perceived value of video games, but not playtime itself: An analysis of Nintendo digital trace data in a casual adult audience"
shorttitle: "Platform-level video gaming and well-being"
author:
  - name: Nick Ballou
    corresponding: false
    orcid: 0000-0003-4126-0696
    email: nick@nickballou.com
    affiliations:
      - name: University of Oxford
        department: Oxford Internet Institute
    roles:
      - conceptualization
      - data curation
      - methodology
      - formal analysis
      - writing

@@

  - name: Andrew K Przybylski
    corresponding: true
    email: andy.pryzbylski@oii.ox.ac.uk
    orcid: 0000-0003-4126-0696
    affiliations:
      - name: University of Oxford
        department: Oxford Internet Institute
    roles:
      - conceptualization
      - funding acquisition
      - project administration
      - editing
abstract: "Studies on video games and well-being often rely on self-report measures or data from a single game. This study of US adults (n = 648) recruited via Prolific analyzes the impact of time spent on over 150 1st-party Nintendo Switch games (totaling 140k hours) on life satisfaction, affective valence, depressive symptoms, and general mental well-being. We replicate previous findings that playtime over the past 2 weeks does not predict well-being, and extend these null findings to a wider range of timescales (1 hour to 1 year). Results suggest that relationships, if present, dissipate within 2 hours of gameplay. Although playtime was not related to well-being, players' assessments of the value of game time---so called gaming life fit---is moderately linked. Our non-causal findings suggest substantial confounding would be needed to shift a meaningful effect to null. Results emphasize the importance of defining the gaming population of interest, collecting data from more than one game, and focusing on how players integrate gaming into their lives rather than the amount of time spent."
keywords: [video games, digital trace data, well-being]
author-note:
  # Disclosures condensed to one paragraph, but you can start a field with two line breaks to break them up: \n\nNew Paragraph
  disclosures:
    study-registration: This study was registered on the Open Science Framework (https://osf.io/sjqyt)
    # Acknowledge and cite data/materials to be shared.
    data-sharing: Data and materials are available at XXXXX. 
    # Example: [Author name] has been a paid consultant for Corporation X, which funded this study.
    conflict-of-interest: The authors declare no competing interests. 
    # Example: This study was supported by Grant [Grant Number] from [Funding Source].
    financial-support: ~ # This research was supported by Huo Family Foundation and the UK Economic and Social Research Council (ESRC) (ES/W012626/1). KM was supported by Forskningsrå det för hälsa, arbetsliv och välfärd (2021-01284).
    # Example: The authors are grateful to [Person] for [Reason].
    gratitude: ~
    # Example. Because the authors are equal contributors, order of authorship was determined by a fair coin toss.
    authorship-agreements: ~
bibliography: references.bib
floatsintext: true # If true, tables and figures are mingled with the text instead of listed at the end of the document.
numbered-lines: false # Numbered lines (.pdf and .docx only)
meta-analysis: false
format:
  apaquarto-docx: default
  # apaquarto-html:
  #   documentmode: man
  #   code-fold: true
  #   code-summary: "Show the code"
  #   wordcount-html: default
  # apaquarto-pdf:
  #   documentmode: jou
---

@@

library(tidyverse)
library(lubridate)
library(knitr)
library(kableExtra)
library(dtplyr)
library(scales)
library(ggdist)
library(distributional)
library(mgcv)
library(see)
library(performance)
library(careless)
library(cowplot)
library(marginaleffects)
library(flextable)

@@

mytheme <- list(
  theme_bw() +
    theme(
      strip.background = element_rect(fill = "black"),
      strip.text = element_text(color = "white", size = 16),
      panel.grid.minor = element_blank(),
      panel.border = element_rect(colour = "black", fill=NA, size=1),
      axis.text.y = element_text(size = 12, hjust = .5, vjust = -.015),
      axis.text.x = element_text(size = 12),
      axis.title.x = element_text(hjust = .45, size = 14, face = "bold"),
      axis.title.y = element_text(hjust = .45, size = 14, face = "bold"),
      legend.direction = "vertical",
      legend.text = element_text(size = 12)
    ),
  coord_cartesian(clip = "off"),
  guides(alpha = "none", fill = guide_legend(reverse = TRUE))
)

@@

Only a handful of studies have applied digital trace data to the study of games and well-being [cf. @VuorreEtAl2022Time; @JohannesEtAl2021video; @BruhlmannEtAl2020Motivational; @LarrieuEtAl2023How], in part because this data can be very difficult to acquire: researchers must build or rely on (often unstable) technical systems to log data themselves, or negotiate individual agreements with games companies who have historically been reluctant to share data [@Ballou2023Manifesto; @SeifEl-NasrEtAl2013game]. Where digital trace data has been collected, however, results have been informative. @BruhlmannEtAl2020Motivational used playtime and in-game behavior to identify subgroups of *League of Legends* players who had more negative in-game experiences. @JohannesEtAl2021video look at playtime in *Animal Crossing: New Horizons* and *PvZ: Battle for Neighborville* and found a positive but likely negligible correlation. A follow-up study expanded this to 7 games, finding that changes in playtime over the course of 6 weeks were unlikely to affect subsequent well-being [@VuorreEtAl2022Time]. @LarrieuEtAl2023How follow high-intensity *Rainbow 6: Siege* players and find no link between playtime and quality of life across any identified player types.

@@

We were also interested in understanding how different timescales relating to play and well-being might be linked. To this end, our second hypothesis focused on how different time lags between when someone plays and is asked about well-being influence the strength of their association. While preregistered, the large number of models implicated in our analysis plan is prohibitively large to be aptly framed as a narrow hypothesis test. We therefore operationalize this as an exploratory research question: How does the relationship between playtime and well-being differ across different gaming observation windows ranging from 1 hour to 1 year?

@@

## Design

Our participant flow is shown in @fig-sankey. We recruited participants from Prolific, a participant recruitment platform, who were: (1) 18 years old, (2) proficient in English, (3) residents of the US, and (4) active video game players (self-defined, using Prolific's built-in screening item). We first distributed a screening questionnaire to 7649 participants asking which video game platforms they were active on; of these, 4184 indicated that they played games on Nintendo Switch.

We invited these participants to a second survey wherein they retrieved an account identifier from the Nintendo web interface (the events QR code, available at <https://accounts.nintendo.com/qrcode>), and shared it with us. This identifier is separate from their username, and cannot be used by anyone besides Nintendo (including ourselves) to link the player to personally identifiable information. A total of 1823 participants undertook and successfully completed the linking process. We sent the account identifiers to Nintendo of America, who in turn sent us each player's pseudonymized play history from May 1, 2022 to present.

@@

dat |> 
  select(age, gender) |> 
  mutate(
    Gender = ifelse(!(gender %in% c("Man","Woman")), 
                     "Non-binary or other gender identity", 
                     gender)
  ) |> 
  group_by(Gender) |> 
  summarise(
    n = n(),
            `Median age (SD)` = paste0(median(age, na.rm = TRUE), " (", round(sd(age, na.rm = TRUE), 2), ")")
    ) |> 
  flextable(cwidth = c(2.5, 1, 1)) %>% 
  theme_apa() 
```

@@

**Video game playtime** was measured by collecting a record of each player's individual game sessions for all 1st party video games played on a Nintendo Switch. These data were provided by Nintendo of America. Playtime was calculated by summing the duration of all (partial) sessions that occur during a given time period relative to the participant's survey response, based on the logged session start and end times. For ease of interpretation, all observation periods longer 24 hours are divided by the number of days in the period, and reported as mean minutes of play per day. It is important to note that these data included only telemetry for games published by Nintendo and its close business partners (e.g. The Pokémon Company), but not games made by third party publishers (e.g. Electronic Arts). Nintendo-published games accounted for 63% of all playtime across our sample; the 37% of play data from 3rd-party games is therefore treated as missing. We return to this limitation in the discussion.

@@

To analyze H1, we fit multiple regression models with playtime over the previous 2 weeks as the primary predictor, all demographic variables as covariates (age, gender, highest level of education, and employment status), and the corresponding mental health variable as the outcome. For example, for H1a (life satisfaction predicted by the previous 2 weeks of play), the model in R is:

{{lm(lifeSat \~ playtimePrev2Weeks + age + gender + education + employment)}}

@@

@@

Given the lack of holistic or platform-level data available in the literature, our first goal was to simply describe the volume of play. This is visualized in @fig-playtime, which show that despite a total play volume of more than 140k hours, our sample was largely minimally engaged with 1st party Nintendo games. During the 2 weeks prior to survey completion, 40% of the sample had no sessions logged at all. The top 10% of players were highly engaged, playing more than 60 minutes per day on average. Sessions of a game lasted on average `r round(median(datRaw$duration, na.rm = T), 1)` minutes \[10th percentile: `r quantile(datRaw$duration, .1)`; 90th percentile: `r quantile(datRaw$duration, .9)`\].

@@

dat |> 
  pivot_longer(cols = c("csas", "affectiveValence", "meanPromis", "meanWemwbs")) |>
  mutate(name = case_when(
    name == "csas" ~ "Life Satisfaction",
    name == "affectiveValence" ~ "Affect",
    name == "meanPromis" ~ "Depressive Symptoms",
    name == "meanWemwbs" ~ "General well-being",
    TRUE ~ NA_character_
    )) |> 
  ggplot(aes(x = playtime2wk*60, y = value)) +
  geom_point(size = .6) +
  geom_smooth(method='lm', formula = y~x, level = .99, color = "darkred") +
  scale_x_sqrt(breaks = c(0, 15, 60, 120, 240, 420)) +
  labs(x = "Minutes of play per day\nduring previous 2 weeks (Square Root Axis)", y = "Well-being (1-5 scale)") +
  mytheme + 
  # scale_y_continuous(labels = scales::percent) +
  facet_wrap(~name)
# plot_predictions(h1csas, condition = "playtime2wk") # test with marginaleffects
```

@@

# fittedModels |> 
#   mutate(
#     model = paste0(outcome, " ~ ", predictor),
#     predictor = fct_rev(fct_recode(predictor, !!!playtimeDict)),
#     outcome = fct_recode(outcome,
#                          "Life Satisfaction" = "csas",
#                          "Affect" = "affectiveValence",
#                          "Depressive Symptoms" = "meanPromis",
#                          "General well-being" = "meanWemwbs"),
#   ) |> 
#   ggplot(
#     aes(y = predictor,
#         fill = predictor,
#         xdist = dist_normal(estimate, std.error))
#   ) +
#   geom_vline(xintercept = 0, color = "grey") +
#   # stat_halfeye(position = position_dodge(.5), # distance between slabs within a group
#   #              normalize = "groups",
#   #              .width = .99, # width of pointrange conf int
#   #              scale = .7) + # height of density plots
#   stat_pointinterval() + 
#   geom_vline(xintercept = -.06, linetype = "dashed") +
#   geom_vline(xintercept = .06, linetype = "dashed") +
#   scale_x_continuous(limits = c(-1, 1), breaks = seq(-1, 1, .5)) +
#   scale_fill_viridis_d(option = "magma", begin = .8, end = .4) +
#   labs(x = "Difference in well-being (1-5 scale)\nAssociated with 1 Additional Hour of Daily Playtime",
#        y = "Playtime Window") +
#   geom_text(aes(x = -.4,
#                 label = str_glue("{number(estimate, .01)} [{number(conf.low, .01)}, {number(conf.high, .01)}]")),
#             vjust = -.3,
#             size = 3,
#             hjust = 1,
#             position = position_dodge(.5)) +
#   # annotate(x = .06, y = +Inf, label = "SESOI", vjust = 2, geom = "label") +
#   # annotate(x = -.06, y = +Inf, label = "SESOI", vjust = 2, geom = "label") +
#   mytheme +
#   theme(legend.position = "none",
#         axis.text.y = element_text(hjust = 1),
#         axis.text.x = element_text(size = 12, hjust = 1)) +
#   facet_wrap(~outcome)

@@

    # some hacky code to match up the time periods with their lubridate::period objects and then convert to numeric
    # so we can plot on the log scale
    window = map(sub("playtime", "", term), ~ timeWindows[.]),
    window = map_dbl(window, ~ as.numeric(.x[[1]], "hours")),
    conf.high.90 = estimate + 1.645*std.error,
    conf.low.90 = estimate - 1.645*std.error
  ) |>
  ggplot(aes(x = window, y = estimate)) +
  geom_hline(
    yintercept = 0,
    linewidth = 0.1
  ) +
  geom_ribbon(
    aes(ymin = conf.low, ymax = conf.high),
    group = 1,
    alpha = .2,
    fill = "darkblue"
  ) +
  geom_ribbon(
    aes(ymin = conf.low.90, ymax = conf.high.90),
    group = 1,
    alpha = .4,
    fill = "darkblue"
  ) +
  geom_line(
    group = 1,
    linewidth = 0.5,
  ) +
  geom_point() +
  scale_x_log10(breaks = map_dbl(timeWindows, ~ as.numeric(.x[[1]], "hours")), labels = names(playtimeDict)) +
  labs(
    x = "Observation period for video game play",
    y = "Association between\noutcome and playtime\nin observation period"
  ) +
  theme_linedraw() +
  theme(axis.title.y = element_text(angle = 0, vjust = .5),
        panel.grid.minor = element_blank()) +
  geom_hline(yintercept = .06, linetype = "dashed", linewidth = .3, color = "black") +
  geom_hline(yintercept = -.06, linetype = "dashed", linewidth = .3, color = "black") +
  facet_wrap(~outcome, scales = "free_y")

@@

We performed various sensitivity checks to ensure the robustness of our findings (see supplementary materials for full detail). First, we explored potential non-linearity in the relationships between playtime and mental health by comparing a generalized additive model of well-being with and without a smooth term for playtime, and comparing AIC between these. Of the 48 possible models (4 well-being variables \* 12 playtime windows), just 1 of these (playtime in the previous 1 year and life satisfaction) showed a difference in AIC of more than 2, indicating that nearly all relationships were adequately captured by linear terms. Next, we reran the analyses using session durations as calculated by Nintendo, as opposed to the implied duration based on the start and end timestamps; Nintendo's durations differ from the implied duration in approximately 10% of sessions, reflecting certain unknown heuristics used by Nintendo to calculate playtime (based e.g., on idle time, time spent in menus, and so on). Data show a similar pattern: no models showed a significant relationship between playtime and well-being at our specified alpha of .01. Next, given that we have relatively sparse data for short playtime windows (e.g., only `r dat |> filter(playtime2hr > 0) |> nrow()` participants had played a session in the 2 hours prior to survey completion), we explored alternative models wherein playtime was separated into both a binary variable (1 if the player had any time logged in that period, 0 if not), and a continuous variable (how much a person played). Results were comparable; although three models indicated that among those who played in the previous 1-2 weeks, longer play is associated with higher affect and general mental well-being, the remaining 93 playtime variables were neither significant or nor within both equivalence bounds.

@@

The question when and how video games cause changes in well-being is at the forefront the minds of researchers, policymakers, parents, and players themselves. Our correlational study cannot claim to answer this causal question as we would first have to identify, to the greatest extent possible, the complete set of confounding factors that jointly cause both play behavior and mental health. Nonetheless, our study is a concrete step in the right direction, having independently recruited a large sample of video game players [as opposed relying on recruitment through games companies themselves; e.g. @JohannesEtAl2021video], collected validated measures of well-being, joined these with objective behavioral telemetry and made minimal adjustments for age, gender, employment, and education. Broadly speaking, we find no evidence to support the existence of a meaningful population-level effect of playtime on well-being.

Although it is not a positive finding, it is informative. This evidence that platform-wide video game play and participant well-being are not meaningfully related across a wide range of outcomes, timescales and model specifications adds to a growing body of work that suggests that there is a low likelihood that population-level affects based on time alone exist in the way that many have historically framed. Said differently, the findings we report place the onus on those who assert that there is a meaningful relationship between playtime and well-being. It should be a priority to identify and concretely articulate which confounds that might bias a true effect towards the robust null associations reported in this and other research using player telemetry [@JohannesEtAl2021video; @VuorreEtAl2022Time; @LarrieuEtAl2023How; @BallouEtAl2024Registered].

To further elucidate this point, we conducted brief simulation tests to ascertain how strong such confounding might need to be (see Supplementary Materials). For example, if the true standardized effect of playtime on mental health was a moderate .2 SDs per additional hour of daily playtime, a confound C would need to be a very strong cause of both X ($beta = .5$) and Y ($beta = -.5$ ) to bias the true .2 effect to null. While we do not claim this is impossible, we do believe it unlikely. Approaching the topic along these lines---identifying confounds, testing the presence or absence of correlations for their sensitivity to potential confounds, and systematically identifying factors that do (not) cause playtime and well-being---can help us achieve more systematic progress. This work can be bolstered by qualitative research aimed at more fully mapping the causal system and by substantive theory development with greater specificity in the aspects of media use expected to produce effects, the hypothesized causal relationships, boundary conditions, and so forth [@MagnussonEtAl2023Harmful; @Ballou2023Manifesto; @Coenen2023Lost; @EronenBringmann2021theory].

@@

The steady attrition throughout the stages of the research process from screening, to linking, to successful data retrieval highlights the challenges for participant recruitment in video games research. Despite a series of filtering steps wherein a majority of participants were filtered out due to not playing Switch games or being unwilling or unable to link data, our final sample remains only minimally engaged with Nintendo games—playing just `r 7*round(median(dat$playtime6mo, na.rm = TRUE), 0)` per week on average. As a result, the population studied here is very different than previous studies whereby players were recruited through targeted emails from games companies [@JohannesEtAl2021video; @LarrieuEtAl2023How] or social media forums for highly-engaged players [@BallouEtAl2024Registered].

As argued above, we believe this is a valuable group in its own right—people who rarely play video games may be particularly susceptible to their positive or negative effects on the rare occasions that they do play. While the current study is unlikely to generalize to so-called "hardcore" players, who may play several hours per day or more and therefore may experience more accumulative effects, our findings align with previous studies of much more highly engaged players and add a new subgroup of players to the body of work showing the absence of any meaningful relationship [@BallouEtAl2024Registered; @LarrieuEtAl2023How; @VuorreEtAl2022Time].

>>>>

---
title: "Perceived value of video games, but not hours played, predicts mental well-being in adult Nintendo players"
shorttitle: "Nintendo gaming and well-being"
author:
  - name: Nick Ballou
    corresponding: false
    orcid: 0000-0003-4126-0696
    email: nick@nickballou.com
    affiliations:
      - name: University of Oxford
        department: Oxford Internet Institute
    roles:
      - conceptualization
      - data curation
      - methodology
      - formal analysis
      - writing

@@

  - name: Andrew K Przybylski
    corresponding: true
    email: andy.pryzbylski@oii.ox.ac.uk
    orcid: 0000-0003-4126-0696
    affiliations:
      - name: University of Oxford
        department: Oxford Internet Institute
    roles:
      - conceptualization
      - funding acquisition
      - project administration
      - editing
abstract: "Studies on video games and well-being often rely on self-report measures or data from a single game. Here, we study how 704 US adults' time spent playing over 150 1st-party Nintendo Switch games (totaling 140k hours) relates to their life satisfaction, affect, depressive symptoms, and general mental well-being. We replicate previous findings that playtime over the past 2 weeks does not predict well-being, and extend these findings to a wider range of timescales (1 hour to 1 year). Results suggest that relationships, if present, dissipate within 2 hours of gameplay. Our non-causal findings suggest substantial confounding would be needed to shift a meaningful true effect to the observed null. Although playtime was not related to well-being, players’ assessments of the value of game time—so called gaming life fit—was. Results emphasize the importance of defining the gaming population of interest, collecting data from more than one game, and focusing on how players integrate gaming into their lives rather than the amount of time spent."
keywords: [video games, digital trace data, well-being]
author-note:
  # Disclosures condensed to one paragraph, but you can start a field with two line breaks to break them up: \n\nNew Paragraph
  disclosures:
    study-registration: This study was registered on the Open Science Framework (https://osf.io/sjqyt).
    # Acknowledge and cite data/materials to be shared.
    data-sharing: Data and materials are available at XXXXX. 
    # Example: [Author name] has been a paid consultant for Corporation X, which funded this study.
    conflict-of-interest: Data for this study was provided by Nintendo of America. Nintendo of America had no role in the design of the study, the analysis of the data, or decision to publish any results. The authors declare no other competing interests. 
    # Example: This study was supported by Grant [Grant Number] from [Funding Source].
    financial-support: ~ This research was supported by Huo Family Foundation and the UK Economic and Social Research Council (ESRC) (ES/W012626/1). KM was supported by Forskningsrå det för hälsa, arbetsliv och välfärd (2021-01284).
    # Example: The authors are grateful to [Person] for [Reason].
    gratitude: ~
    # Example. Because the authors are equal contributors, order of authorship was determined by a fair coin toss.
    authorship-agreements: ~
bibliography: references.bib
floatsintext: true # If true, tables and figures are mingled with the text instead of listed at the end of the document.
numbered-lines: false # Numbered lines (.pdf and .docx only)
meta-analysis: false
format:
  apaquarto-docx: default
  # apaquarto-html:
  #   documentmode: man
  #   code-fold: true
  #   code-summary: "Show the code"
  #   wordcount-html: default
  # apaquarto-pdf:
  #   documentmode: jou
---

@@

library(tidyverse)
library(lubridate)
library(knitr)
library(kableExtra)
library(dtplyr)
library(scales)
library(ggdist)
library(distributional)
library(mgcv)
library(patchwork)
library(see)
library(performance)
library(careless)
library(cowplot)
library(marginaleffects)
library(flextable)

@@

mytheme <- list(
  theme_bw() +
    theme(
      strip.background = element_rect(fill = "black"),
      strip.text = element_text(color = "white", size = 10),
      panel.grid.minor = element_blank(),
      panel.border = element_rect(colour = "black", fill=NA, size=1),
      axis.text.y = element_text(size = 10, hjust = .5, vjust = -.015),
      axis.text.x = element_text(size = 10),
      axis.title.x = element_text(hjust = .45, size = 10, face = "bold"),
      axis.title.y = element_text(hjust = .45, size = 10, face = "bold"),
      legend.direction = "vertical",
      legend.text = element_text(size = 10)
    ),
  # coord_cartesian(clip = "off"),
  guides(alpha = "none", fill = guide_legend(reverse = TRUE))
)

@@

Only a handful of studies have applied digital trace data to the study of games and well-being \[ @VuorreEtAl2022Time; @JohannesEtAl2021video; @BruhlmannEtAl2020Motivational; @LarrieuEtAl2023How\], in part because this data can be very difficult to acquire: researchers must build or rely on (often unstable) technical systems to log data themselves, or negotiate individual agreements with games companies who have historically been reluctant to share data [@Ballou2023Manifesto; @SeifEl-NasrEtAl2013game]. Where digital trace data has been collected, however, results have been informative. @BruhlmannEtAl2020Motivational used playtime and in-game behavior to identify subgroups of *League of Legends* players who had more negative in-game experiences. @JohannesEtAl2021video look at playtime in *Animal Crossing: New Horizons* and *PvZ: Battle for Neighborville* and found a positive but likely negligible correlation. A follow-up study expanded this to 7 games, finding that changes in playtime over the course of 6 weeks were unlikely to affect subsequent well-being [@VuorreEtAl2022Time]. @LarrieuEtAl2023How follow high-intensity *Rainbow 6: Siege* players and find no link between playtime and quality of life across any identified player types.

@@

We were also interested in understanding how relationships between play and well-being might depend on the choice of what time scale of play is considered relevant. To this end, we examined relationships between well-being and a wide range of time windows of play preceding the well-being question. While preregistered, the large number of models implicated in our analysis plan is prohibitively large to be aptly framed as a narrow hypothesis test. We therefore operationalize this as an exploratory research question: How does the relationship between playtime and well-being differ across different gaming observation windows ranging from 1 hour to 1 year?

@@

## Design and Recruitment

Our participant flow is shown in @fig-sankey. We recruited participants from Prolific, a participant recruitment platform, who were: (1) 18 years old, (2) proficient in English, (3) residents of the US, and (4) active video game players (self-defined, based on a \>0 response to Prolific's built-in screening item "How many hours per week do you play video games on average?"). We first distributed a screening questionnaire to 7649 participants asking which video game platforms they were active on; of these, 4184 indicated that they played games on Nintendo Switch.

We invited these participants to a second survey wherein they retrieved an account identifier from the Nintendo web interface (the events QR code, available at <https://accounts.nintendo.com/qrcode>), and shared it with us. This identifier is separate from their username, and cannot be used by anyone besides Nintendo (including ourselves) to link the player to personally identifiable information. A total of 1823 participants completed the linking process. We sent the account identifiers to Nintendo of America, who in turn sent us each player's pseudonymized play history from May 1, 2022 to present. Data collection began with a pilot study of 100 participants on Nov 15, 2023, which was combined with primary data collection from Feb 12, 2024 to May 6, 2024.

@@

dat |> 
  select(age, gender) |> 
  mutate(
    Gender = ifelse(!(gender %in% c("Man","Woman")), 
                     "Non-binary or other gender identity", 
                     gender)
  ) |> 
  group_by(Gender) |> 
  summarise(
    n = n(),
            `Median age (SD)` = paste0(median(age, na.rm = TRUE), " (", round(sd(age, na.rm = TRUE), 2), ")")
    ) |> 
  flextable(cwidth = c(2.5, 1, 1)) %>% 
  theme_apa()

@@

**Video game playtime** was measured by collecting a record of each player's individual game sessions for all 1st party video games played on a Nintendo Switch. These data were provided by Nintendo of America. Playtime was calculated by summing the duration of all (partial) sessions that occur during a given time period relative to the participant's survey response, based on the logged session start and end times. For ease of interpretation, game play time in all observation periods longer 24 hours is reported as mean minutes of play per day. It is important to note that these data included only telemetry for games published by Nintendo and its close business partners (e.g. The Pokémon Company), but not games made by third party publishers (e.g. Electronic Arts). Nintendo-published games accounted for 63% of all playtime across our sample; the 37% of play data from 3rd-party games is therefore treated as missing. We return to this limitation in the discussion.

@@

To test H1, we fit multiple regression models with playtime over the previous 2 weeks as the primary predictor, all demographic variables as covariates (age, gender, highest level of education, and employment status), and the corresponding mental health variable as the outcome. For example, for H1a (life satisfaction predicted by the previous 2 weeks of play), the model in R is:

{{lm((life satisfaction) \~ (playtime in the previous 2 weeks) + age + gender + education + employment)}}

@@

dat |> 
  select(playtime1d, playtime2wk, playtime3mo, playtime1yr) |> 
  # convert back to total playtime in hours, rather than minutes per day
  # mutate(
  #   playtime1d = playtime1d*24,
  #   playtime2wk = playtime2wk*24*14,
  #   playtime3mo = playtime3mo*24*91,
  #   playtime1yr = playtime1yr*24*36
  # ) |> 
  # scale_x_continuous(labels = scales::cut_si()) +
  pivot_longer(cols = everything(), 
               names_to = "timewindow", 
               names_pattern = "playtime(.*)", 
               values_to = "playtime") |> 
  mutate(timewindow = fct_recode(timewindow,
                                 "1 Day" = "1d",
                                 "2 Weeks" = "2wk",
                                 "3 Months" = "3mo",
                                 "1 Year" = "1yr"),
         timewindow = fct_relevel(timewindow, 
                                  "1 Day", 
                                  "2 Weeks", 
                                  "3 Months", 
                                  "1 Year"),
         # playtime = if_else(playtime == 0, map_dbl(playtime, ~ sample(c(0, -.1, -.2, -.3), 1)), playtime)
  ) |> 
  ggplot(aes(x = timewindow, y = playtime, fill = timewindow)) +
  stat_slab(
    density = ggdist::density_bounded(bandwidth = "nrd0"),
    normalize = "all",
    # height = 2
    scale = .4,
  ) +
  stat_dots(
    side = "left",
    justification = 1.2,
    overflow = "compress",
    # overflow = "multiple",
    binwidth = .1,
    dotsize = .5,
    scale = .4,
    color = NA,
    layout = "hex",
    # stackratio = .5,
    overlaps = "nudge",
    # verbose = TRUE,
    alpha = .3
  ) +
  geom_boxplot(
    width = .10, 
    outlier.shape = NA
  ) +
  labs(x = "Observation Period", y = "Daily Hours of Playtime") +
  scale_y_continuous(
    trans = "sqrt",
    breaks = c(0, .2, .5, 1, 2, 3, 4, 5),
    limits = c(0, 5)
    # limits = c(-.3, 5),
  ) +
  scale_x_discrete(expand = c(0, 0)) +
  coord_cartesian(xlim = c(.3, 4.3)) +
  mytheme +
  theme(legend.position = "none") +
  scale_fill_viridis_d()

@@

Given the lack of holistic or platform-level data available in the literature, our first goal was to simply describe the volume of play. This is visualized in @fig-playtime, which show that despite a total play volume of more than 140k hours, our sample was largely minimally engaged with 1st party Nintendo games. During the 2 weeks prior to survey completion, just over half of the sample had 0 sessions logged. The top 10% of players were moderately engaged, playing an average of 60 minutes per day. Sessions of a game lasted on average `r round(median(datRaw$duration, na.rm = T), 1)` minutes \[10th percentile: `r quantile(datRaw$duration, .1)`; 90th percentile: `r quantile(datRaw$duration, .9)`\].

@@

dat |> 
  pivot_longer(cols = c("csas", "affectiveValence", "meanPromis", "meanWemwbs")) |>
  mutate(name = case_when(
    name == "csas" ~ "Life Satisfaction",
    name == "affectiveValence" ~ "Affect",
    name == "meanPromis" ~ "Depressive Symptoms",
    name == "meanWemwbs" ~ "General well-being",
    TRUE ~ NA_character_
    )) |> 
  ggplot(aes(x = playtime2wk*60, y = value)) +
  geom_point(size = .6) +
  geom_smooth(method='lm', formula = y~x, level = .99, color = "darkgreen", fill = "darkgrey") +
  scale_x_sqrt(breaks = c(0, 15, 60, 120, 240, 420)) +
  labs(x = "Minutes of play per day\nduring previous 2 weeks (Square Root Axis)", y = "Outcome (1-5 scale)") +
  mytheme +
  # theme(
  #   axis.line.y = element_line(linewidth = .5)
  # ) + 
  # scale_y_continuous(labels = scales::percent) +
  facet_wrap(~name)

# plot_predictions(h1csas, condition = "playtime2wk") # test with marginaleffects
```

@@

@@

    # some hacky code to match up the time periods with their lubridate::period objects and then convert to numeric
    # so we can plot on the log scale
    window = map(sub("playtime", "", term), ~ timeWindows[.]),
    window = map_dbl(window, ~ as.numeric(.x[[1]], "hours")),
    conf.high.90 = estimate + 1.645*std.error,
    conf.low.90 = estimate - 1.645*std.error
  ) |>
  ggplot(aes(x = window, y = estimate)) +
  geom_ribbon(
    aes(ymin = conf.low, ymax = conf.high),
    group = 1,
    alpha = .4,
    fill = "darkgrey"
  ) +
  geom_ribbon(
    aes(ymin = conf.low.90, ymax = conf.high.90),
    group = 1,
    alpha = .8,
    fill = "darkgrey"
  ) +
  geom_line(
    group = 1,
    color = "darkgreen",
    linewidth = 1,
  ) +
  geom_point(color = "darkgreen", size = 2) +
  scale_x_log10(breaks = map_dbl(timeWindows, ~ as.numeric(.x[[1]], "hours")), labels = names(playtimeDict)) +
  labs(
    x = "Observation period for video game play",
    y = "Playtime-outcome association\nin observation period"
  ) +
  theme_linedraw() +
  theme(
    # axis.title.y = element_text(angle = 0, vjust = .5),
    panel.grid.minor = element_blank(),
    panel.grid.major = element_line(linewidth = .1)
    ) +
  geom_hline(yintercept = .06, linetype = "dashed", linewidth = .3, color = "black") +
  geom_hline(yintercept = -.06, linetype = "dashed", linewidth = .3, color = "black") +
  facet_wrap(~outcome, scales = "free_y")

@@

We performed various sensitivity checks to ensure the robustness of our findings (see supplementary materials for full detail). First, we explored potential non-linearity in the relationships between playtime and mental health by comparing a generalized additive model of well-being with and without a smooth term for playtime, and comparing AIC between these. Of the 48 possible models (4 well-being variables \* 12 playtime windows), just 1 of these (playtime in the previous 1 year and life satisfaction) showed a difference in AIC of more than 2, indicating that nearly all relationships were adequately captured by linear terms. Next, we reran the analyses using session durations as calculated by Nintendo, as opposed to the implied duration based on the start and end timestamps; Nintendo's durations differ from the implied duration in approximately 10% of sessions, reflecting certain unknown heuristics used by Nintendo to calculate playtime (based e.g., on idle time, time spent in menus, and so on). Data show a similar pattern: no models showed a significant relationship between playtime and well-being at our specified alpha of .01. Next, we explored alternative models wherein playtime was separated into both a binary variable (1 if the player had any time logged in that period, 0 if not), and a continuous variable (how much a person played). Results were comparable; although three models indicated that among those who played in the previous 1-2 weeks, longer play is associated with higher affect and general mental well-being, the remaining 93 playtime variables were neither significant or nor within both equivalence bounds.

@@

Although we did not design our study to test the causal question most critical to stakeholders, our study is a concrete step in the right direction, having independently recruited a large sample of video game players [as opposed relying on recruitment through games companies themselves; e.g. @JohannesEtAl2021video], collected validated measures of well-being, joined these with objective behavioral telemetry and made minimal adjustments for age, gender, employment, and education. Using these methods, we did not find robust or consistent relationships between time spent playing and various mental well-being outcomes.

Although not conclusive, our results point toward a pattern whereby platform-wide video game play time does not predict well-being to a meaningful degree. This trend, across a wide range of outcomes, timescales of play, and model specifications adds to a growing body of work that suggests that simple time spent playing games is unlikely to affect well-being for the average player. Said differently, the findings we report place the onus on those who assert that there is a meaningful relationship between playtime and well-being. It should be a priority to identify and concretely articulate which confounds might bias a true effect towards the null associations reported in this and other research using player telemetry [@JohannesEtAl2021video; @VuorreEtAl2022Time; @LarrieuEtAl2023How; @BallouEtAl2024Registered].

To further elucidate this point, we conducted brief simulation tests to ascertain how strong such confounding might need to be (see Supplementary Materials). For example, if the true standardized effect of playtime on mental health was a moderate .2 SDs per additional hour of daily playtime, a confound C would need to be a very strong cause of both X ($beta = .5$) and Y ($beta = -.5$ ) to bias the true .2 effect to null. While we do not claim this is impossible, we do believe it unlikely. Approaching the topic along these lines---identifying confounds, testing the presence or absence of correlations for their sensitivity to potential confounds, and systematically identifying factors that do (not) cause playtime and well-being---can help us achieve more systematic progress [@BallouEtAl2024How]. This work can be bolstered by qualitative research aimed at more fully mapping the causal system and by substantive theory development with greater specificity in the aspects of media use expected to produce effects, the hypothesized causal relationships, boundary conditions, and so forth [@MagnussonEtAl2024Harmful; @Ballou2023Manifesto; @Coenen2023Lost; @EronenBringmann2021theory].

@@

The steady attrition throughout the stages of the research process from screening, to linking, to successful data retrieval highlights the challenges for participant recruitment in video games research. Despite a series of filtering steps wherein a majority of participants were filtered out due to not playing Switch games or being unwilling or unable to link data, our final sample remains only minimally engaged with Nintendo games—playing just `r 7*round(median(dat$playtime1yr, na.rm = TRUE), 1)` hours per week on average. As a result, the population here is very different than previous studies whereby players were recruited through targeted emails from games companies [@JohannesEtAl2021video; @LarrieuEtAl2023How] or social media forums for highly-engaged players [@BallouEtAl2024Registered].

As argued above, we believe this is a valuable group in its own right—people who rarely play video games may be particularly susceptible to their positive or negative effects on the occasions they do play. While the current study is unlikely to generalize to so-called "hardcore" players who play several hours per day or more and therefore may experience more accumulative effects, our findings align with previous studies of more highly engaged players and add a new subgroup of players to the body of work showing the absence of any meaningful relationship [@BallouEtAl2024Registered; @LarrieuEtAl2023How; @VuorreEtAl2022Time].
